Ecological performance of detecting data fabrication with summary statistics

### 2016-09-01 submission/manuscript.tex
Update manuscript files

To detect data fabrication in a set of summary results, we first tested the standardized standard deviations (SDs) for data fabrication \citep{simonsohn2013} across the four anchoring studies. Such methods test whether the observed SDs contain a reasonable amount of variation, as expected based on random sampling processes. For example, if four independent samples all yield the variance $2.22$, this could be considered excessively consistent when the probability that this amount of consistency (or more) is less than 1 out of 1000 in truly random samples. To compute this test, we first standardized the SDs for each of the four studies by computing
\begin{equation}
z_j=\sqrt{\frac{s^2_j}{MS_w}}=\sqrt{\frac{s^2_j}{\left(\frac{\sum\limits^k_{j=1}(N_j-1)s^2_j}{\sum\limits^k_{j=1}(N_j-1)}\right)}}
\label{s2_j}
\end{equation}
where $z_j$ denotes the standardized SD in group $j$ ($MS_w$ is the simple arithmetic mean when sample sizes are equal for all cells, which is the case for the fabricated datasets). We tested different measures to detect data fabrication that utilize these standardized SDs (i.e., $z_j$). We included the variance of the standardized SDs \citep[i.e.,  $SD_{z}$;][]{simonsohn2013} and tried out the max-min distance of the standardized SDs (denoted $max-min_{z}$) as an alternative measure. We compared the observed value for each measure with the expected distribution when the summary results are used to generate random samples. To this end, we simulated the expected distribution of standardized SDs and computed the expected distribution of each measure. This expected distribution was used to determine the $p$-value of the observed $SD_z$ and $max-min_z$. We simulated the standardized variance for each of the $j$ groups as
\begin{equation}
z^2_j\sim\left(\frac{\chi^2_{N_j-1}}{N_j-1}\right)/MS_w
\label{simvar}
\end{equation}
These simulated values are used to compute the expected distribution of the $SD_z$ and $max-min_z$  measures.

Second, we applied the reversed Fisher method to the nonsignificant $p$-values twice, once for the gender effects hypothesis and once for the interaction effects hypothesis, in order to detect data fabrication. The Fisher method \citep{fisher1925} tests for evidence of an effect in a set of $p$-values by testing for a right-skew $p$-value distribution, but we adjusted it here to test for results that are overly consistent with the null hypothesis and result in a left-skew distribution (see Figure \ref{leftskew}). The original Fisher method is computed as
\begin{equation}
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(p_i)
\label{fisher}
\end{equation}
and tests for right-skew in a set of $p$-values, but we adjust it to the following
\begin{equation}
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(1-\frac{p_i-t}{1-t})
\label{fisheradjusted}
\end{equation}
where it now tests for left-skew (i.e., more larger $p$-values than smaller $p$-values) across the $k$ number of $p$-values that falls above the threshold $t$. We set this threshold to .05 in order to include only nonsignificant test results.

% add plot of left and right skew
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{../figures/fisherfig.pdf}
\caption{Conceptual representation of what the Fisher test inspects (Equation \ref{fishertest}; panel A) and our adjusted Fisher test inspects (Equation \ref{fisheradjusted}; panel B). Both panels test whether there is sufficient evidence that the solid line deviates from the dashed line, except that the type of deviation that the test is sensitive to is the exact opposite.}
\label{leftskew}
\end{center}
\end{figure}

% Finally, we combined the results of these three individual tests for data fabrication using the Fisher method. We expect this combination test of the three individual tests for data fabrication to be a more powerful than the individual tests. Based on the results of the combined test results, the three 'best' data fabricators are selected. The three respondents with the highest *p*-values contain the least evidential value for deviating from genuine data and receive an additional $50 Amazon gift card.
% 
% For each of these three tests to detect data fabrication we carry out sensitivity and specificity analyses using ROC-curves in order to determine optimal alpha levels. This analysis helps determine the classification performance of these three tests. In order to determine the optimal alpha level, we varied the alpha level from .000001 through .1 and assessed the classification performance of these methods to detect data fabrication. The optimal alpha level is determined by finding that alpha level for which both the true positive classification rate and the true negative classification rate are highest. For example,

@@

\bibliography{../bibliography/library}
\end{document}

>>>>

To detect data fabrication in a set of summary results, we first tested the standardized standard deviations (SDs) for data fabrication \citep{simonsohn2013} across the four anchoring studies. This method tests whether the observed SDs contain a reasonable amount of variation, as expected based on random sampling processes. For example, if four independent samples all yield the variance $2.22$, this could be considered excessively consistent when the probability that this amount of consistency (or more) is less than 1 out of 1000 in truly random samples. To compute this probability, we first standardized the SDs for each of the four studies with
\begin{equation}
z_j=\sqrt{\frac{s^2_j}{MS_w}}=\sqrt{\frac{s^2_j}{\left(\frac{\sum\limits^k_{j=1}(N_j-1)s^2_j}{\sum\limits^k_{j=1}(N_j-1)}\right)}}
\label{s2_j}
\end{equation}
where $z_j$ denotes the standardized SD in group $j$ ($MS_w$ is the simple arithmetic mean when sample sizes are equal for all cells, which is the case for the fabricated datasets). We tested different measures to detect data fabrication that utilize these standardized SDs (i.e., $z_j$). We included the variance of the standardized SDs \citep[i.e.,  $SD_{z}$;][]{simonsohn2013} and tried out the max-min distance of the standardized SDs (denoted $max-min_{z}$) as an alternative measure. We compared the observed value for each measure with the expected distribution when the summary results are used to generate random samples. To this end, we simulated the expected distribution of standardized SDs and computed the expected distribution of each measure. This expected distribution was used to determine the $p$-value of the observed $SD_z$ and $max-min_z$. We simulated the standardized variance for each of the $j$ groups as
\begin{equation}
z^2_j\sim\left(\frac{\chi^2_{N_j-1}}{N_j-1}\right)/MS_w
\label{simvar}
\end{equation}
These simulated values are used to compute the expected distribution of the $SD_z$ and $max-min_z$  measures.

Testing the standardized SDs for potential data fabrication can be done either for each study separately or all studies combined; the test can also be done under different assumptions of population variances across conditions. The assumptions of population variance can either by that all SDs originate from the same distribution \citep[as in][]{simonsohn2013}, the SDs within a factor are from the same distribution, or each group comes from its own distribution. We preregistered the method that assumes the SDs are drawn from the same distribution for the various conditions (i.e., homogeneous SDs) and are tested across all studies. However, upon conducting the analyses, we decided homogenous SDs are not unequivocal and included computations where the SDs for the low anchor and high anchor are from different distributions (i.e., heterogeneous SDs). Additionally, the signal for data fabrication across the four anchoring studies might result in different studies cancelling each other out, so we also included analyses where each study was analyzed separately.

Second, we applied the reversed Fisher method to detect data fabrication to the nonsignificant $p$-values twice: once for the results of gender effects hypothesis in each study and once for the results of the interaction effect hypothesis for each study. The Fisher method \citep{fisher1925} tests for evidence of an effect in a set of $p$-values by testing for a right-skew $p$-value distribution, but we adjusted it here to test for results that are overly consistent with the null hypothesis and result in a left-skew distribution (see Figure \ref{leftskew}). The original Fisher method is computed as
\begin{equation}
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(p_i)
\label{fishertest}
\end{equation}
and tests for right-skew in a set of $p$-values, but we adjust it to the following
\begin{equation}
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(1-\frac{p_i-t}{1-t})
\label{fisheradjusted}
\end{equation}
where it now tests for left-skew (i.e., more larger $p$-values than smaller $p$-values) across the $k$ number of $p$-values that falls above the threshold $t$. We set this threshold to .05 in order to include only nonsignificant test results.

% add plot of left and right skew
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{../figures/fisherfig.pdf}
\caption{Conceptual representation of what the Fisher test inspects (Equation \ref{fishertest}; panel A) and what the adjusted Fisher test inspects (Equation \ref{fisheradjusted}; panel B). Both panels test whether there is sufficient evidence that the solid line deviates from the dashed line, except that the type of deviation that the test is sensitive to is the exact opposite.}
\label{leftskew}
\end{center}
\end{figure}

Finally, we combined the three methods to detect data fabrication into an overall detection method (i.e., combination method). This included the $SD_z$ measure across all studies and the Fisher test (Equation \ref{fisheradjusted}) of the gender hypothesis test and the interaction test.  We expected this combination test of the three individual tests for data fabrication to be more powerful than the individual tests, given that these tests inspect different manifestations of data fabrications. Based on the results of the combined test results, the three least detectable data fabricators were selected. The three respondents with the highest $p$-values on this combined method to detect data fabrication contained the least evidential value for deviating from genuine data and received an additional \$50 Amazon gift card.

For each of these four tests to detect data fabrication ($SD_z$, Fisher test for the gender and interaction hypotheses, combination method) we carried out sensitivity and specificity analyses using ROC-curves. With these ROC-curves, more informed decisions about optimal alpha levels can be made. In order to determine the optimal alpha level, we varied the alpha level from .000001 through .1. The optimal alpha level was determined by finding that alpha level for which the combination of true positive rate (TPR) and the true negative rate (TNR) were highest. For example, if $\alpha=.04$ results in $TPR=.30$ and $TNR=.7$, but $\alpha=.05$ results in $TPR=.5$ and $TNR=.5$, then .05 was chosen as an optimal decision criterion based on the sample.

@@

The collected data included 36 genuine data from Many Labs 1 \citep[\url{osf.io/pqf9r};][]{klein2014} and 39 fabricated datasets(\url{osf.io/e6zys}). 3 participants did not participate for a bonus. Figure \ref{densities} shows a group-level comparison of the genuine- and fabricated $p$-values and effect sizes ($r$). Such group-level comparisons provide a general idea of the differences between the genuine- and fabricated data \citep[see also][]{Akhtar-Danesh2003}.

% add plot of left and right skew
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{../figures/ddfab_density.pdf}
\caption{Overlay of density distributions for both genuine and fabricated data, per effect and type of result. We instructed respondents to fabricate nonsignificant data for the gender and interaction effects, and a significant effect for the condition effect.}
\label{densities}
\end{center}
\end{figure}

The distributions in Figure \ref{densities} indicate little difference between genuine- and fabricated data when nonsignificant effects are inspected (i.e., gender and interaction hypotheses) whereas there seem to be large differences when subjects were required to fabricate significant data (i.e., condition hypothesis). From our own experience, and anecdotal evidence elsewhere \citep{BAILEY1991741}, large effects have raised initial suspicions previously. These data confirm that extremely large effect sizes (e.g., $r>.95$) might prove to be an easy-to-implement flag for potentially anomalous data (it is wise to seek for alternative explanations after flagging, nonetheless). Considering this, we also investigated how well effect sizes perform in detecting data fabrication (not preregistered).

\subsubsection*{Performance of using SDs to detect data fabrication}

% latex table generated in R 3.3.1 by xtable 1.8-2 package
% Thu Sep  1 17:46:56 2016
\begin{table}[ht]
\centering
\begin{tabular}{lrr}
  \hline
Method & $SD_z$ & $maxmin_z$ \\ 
  \hline
Homogeneous combined & 0.423 & 0.303 \\ 
  Homogeneous study 1 & 0.368 & 0.374 \\ 
  Homogeneous study 2 & 0.421 & 0.446 \\ 
  Homogeneous study 3 & 0.510 & 0.521 \\ 
  Homogeneous study 4 & 0.540 & 0.542 \\ 
  Heterogeneous combined & 0.770 & 0.756 \\ 
  Study 1 low condition & 0.643 & 0.643 \\ 
  Study 1 high condition & 0.438 & 0.438 \\ 
  Study 2 low condition & 0.751 & 0.751 \\ 
  Study 2 high condition & 0.612 & 0.612 \\ 
  Study 3 low condition & 0.668 & 0.668 \\ 
  Study 3 high condition & 0.651 & 0.651 \\ 
  Study 4 low condition & 0.796 & 0.796 \\ 
  Study 4 high condition & 0.555 & 0.555 \\ 
   \hline
\end{tabular}
\caption{Area Under the Curve (AUC) for the two statistical methods used to detect data fabrication based on variances. Homogeneous: assumes one population variance underlying all groups. Heterogeneous: assumes separate population variance per anchoring condition.} 
\label{auc_variances}
\end{table}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{../figures/varianceplot.pdf}
\caption{PLACEHOLDER}
\label{varianceplot}
\end{center}
\end{figure}

\subsubsection*{Performance of using $p$-values to detect data fabrication}

We asked researchers to fabricate data for nonsignificant effect sizes after inspecting whether the genuine data resembled a uniform distribution. Consequently, our hypothesis was that researchers who fabricate data would be unlikely to fabricate a uniform $p$-value distribution, given many researchers misinterpret the $p$-value (e.g., as the probability of an effect %REF
). However, as Table \ref{auc_fisher} indicates, the AUC for these methods are hardly better than chance. In other words, for effects that are specifically fabricated to be nonsignificant, researchers might not fabricate such large $p$-values to be suspicious.

% latex table generated in R 3.3.1 by xtable 1.8-2 package
% Thu Sep  1 17:46:56 2016
\begin{table}[ht]
\centering
\begin{tabular}{lr}
  \hline
Method & AUC \\ 
  \hline
Fisher test gender hypothesis & 0.521 \\ 
  Fisher test interaction hypothesis & 0.535 \\ 
   \hline
\end{tabular}
\caption{Area Under the Curve (AUC) for PLACEHOLDER} 
\label{auc_fisher}
\end{table}
\subsubsection*{Performance of combining SDs and $p$-value methods to detect data fabrication}

% latex table generated in R 3.3.1 by xtable 1.8-2 package
% Thu Sep  1 17:46:56 2016
\begin{table}[ht]
\centering
\begin{tabular}{lr}
  \hline
Method & AUC \\ 
  \hline
Combined Fisher test (3 results, $SD_z$ homogeneous) & 0.602 \\ 
  Combined Fisher test (3 results, $SD_z$ heterogeneous) & 0.736 \\ 
  Combined Fisher test (6 results, $SD_z$ homogeneous) & 0.643 \\ 
  Combined Fisher test (6 results, $SD_z$ heterogeneous) & 0.643 \\ 
   \hline
\end{tabular}
\caption{Area Under the Curve (AUC) for PLACEHOLDER} 
\label{auc_combined}
\end{table}

\subsubsection*{Performance of large effect sizes to detect data fabrication}

%The 23 largest significant effects were fabricated, with 13 (0.333\%) larger than $r=.9$. Based on descriptive data throughout psychology \citep{Hartgerink2016-lg}, effect sizes $\geq.9$ belong to the 96th percentile of effects throughout the psychology literature and effect sizes $\geq.95$ 98th percentile.

\subsection*{Discussion}

% variance analysis and effect size best ways to inspect
% but severely correlated: smaller variances = larger effects
% easiest way is too look for massive effects in the effects

% Simonsohn method not invalid under homogeneous, just somethign to take into account
% easily adjusted

% Discuss how the PPV is likely to be overestimated because prevalence is 50% in this study
% How we don't know the prevalence, in fact.

\bibliography{../bibliography/library}
\newpage
%------------------------------------
%handy to include this at the end
%------------------------------------
\section*{SessionInfo}
%-------------------------------------

\begin{Schunk}
\begin{Sinput}
> sessionInfo()
\end{Sinput}
\begin{Soutput}
R version 3.3.1 (2016-06-21)
Platform: x86_64-redhat-linux-gnu (64-bit)
Running under: Fedora 23 (Workstation Edition)

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
 [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
 [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
 [9] LC_ADDRESS=C               LC_TELEPHONE=C            
[11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base

other attached packages:
 [1] stringr_1.0.0   plyr_1.8.4      car_2.0-19      httr_1.2.0     
 [5] xtable_1.8-2    gridExtra_2.2.1 ggplot2_2.1.0   latex2exp_0.4.0
 [9] foreign_0.8-66  pROC_1.8

loaded via a namespace (and not attached):
 [1] Rcpp_0.12.6      MASS_7.3-45      R6_2.1.2         grid_3.3.1      
 [5] gtable_0.2.0     magrittr_1.5     scales_0.4.0     stringi_1.1.1   
 [9] tools_3.3.1      munsell_0.4.3    colorspace_1.2-6 nnet_7.3-12     
\end{Soutput}
\begin{Sinput}
> 
\end{Sinput}
\end{Schunk}

\end{document}
